An algorithm for the optimization of finite element integration loops

### 2016-12-23 TOMS_Paper/assembly.tex
Lots of fixes towards final version

\section{Transformation Space}
\label{sec:optimal-impl}
In this section, we characterize global and local optimality for finite element integration as well as the space of legal transformations that needs be explored to achieve them. The method by which exploration is performed is discussed in Section~\ref{sec:optimal-synthesis}.

@@

\section{Code Generation}
\label{sec:codegen}
Sharing elimination and pre-evaluation, as well as the transformation algorithm, have been implemented in COFFEE, the compiler for finite element integration routines adopted in Firedrake. In this section, we briefly discuss the aspects of the compiler that are relevant for this article.

@@

\section{Performance Evaluation}
\label{sec:perf-results}

@@

The Zenodo system was used to archive all packages used to perform the experiments: Firedrake \cite{lawrence_mitchell_2016_49284}, PETSc \cite{barry_smith_2016_49285}, petsc4py \cite{firedrake_2016_49283}, FIAT \cite{marie_e_rognes_2016_49280}, UFL \cite{martin_sandve_alnaes_2016_49282}, FFC \cite{anders_logg_2016_49276}, PyOP2 \cite{florian_rathgeber_2016_49281} and COFFEE \cite{fabio_luporini_2016_49279}. The experiments can be reproduced using a publicly available benchmark suite~\cite{florian_rathgeber_2016_49290}.

@@

Following the methodology adopted in~\citeN{quadrature1}, we vary the following parameters:
\begin{itemize}
\item the polynomial degree of test, trial, and coefficient (or ``pre-multiplying'') functions, $q \in \lbrace1, 2, 3, 4\rbrace$
\item the number of coefficient functions $\mathrm{nf} \in \lbrace0, 1, 2, 3\rbrace$
\end{itemize}
While constants of our study are
\begin{itemize}
\item the space of test, trial, and coefficient functions: Lagrange
\item the mesh: tetrahedral with a total of 4374 elements
\item exact numerical quadrature (we employ the same scheme used in~\citeN{quadrature1}, based on the Gauss-Legendre-Jacobi rule)
\item double-precision arithmetic
\end{itemize}

@@

\subsection{Autovectorization}

We chose the mesh dimension and the function spaces such that the inner loop sizes would always be a multiple of the machine vector length. This ensured autovectorization in the majority of code variants\footnote{We verified the vectorization of inner loops by looking at both compiler reports and assembly code.}. The biggest exception is \texttt{quad}, due to the presence of indirection arrays in the generated code. In \texttt{tens}, loop nests are fully unrolled, so the standard loop vectorization is not feasible; the compiler reports suggest, however, that block vectorization \cite{SLP-vect} is often triggered. In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have identical structure, with loop vectorization being regularly applied.

@@

We first notice the lack of improvements when $q=1$. The operation count of \texttt{cfO2} and \texttt{tensor} is roughly 9$\times$ lower than that of the alternatives (tens instead of hundreads operations), but nontheless no gain in execution time is obtained. This is due to the fact that local assembly is memory-bound. For instance, in the simplest case $nf=0$, in addition to the determinant of the Jacobian inverse, \texttt{tens} only requries three unique multiplications to evaluate the local element matrix (26 operations in total), whereas almost a hundred memory accesses are performed. In particular, we have: 12 loads for the vertex coordinates; for each vertex, 1 load to indirectly access its data; for each of the 16 local elements, 4 loads (optimistically) and 1 store to update an entry in the global matrix in compressed sparse row format. This unsurprisingly leads to a very low arithmetic intensity (i.e., $\frac{\text{operation count}}{\text{bytes moved}}$), which collocates the kernels in a memory-bound regime.

@@

\begin{figure}
\includegraphics[scale=0.685]{perf-results/elasticity}
\caption{Performance evaluation for the bilinear form arising in an \textit{elastic} model. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
\end{figure}

@@

@@

>>>>

\section{Transformation space}
\label{sec:optimal-impl}
In this section, we characterize global and local optimality for finite element integration as well as the space of legal transformations that needs be explored to achieve them. The method by which exploration is performed is discussed in Section~\ref{sec:optimal-synthesis}.

@@

\section{Code generation}
\label{sec:codegen}
Sharing elimination and pre-evaluation, as well as the transformation algorithm, have been implemented in COFFEE, the compiler for finite element integration routines adopted in Firedrake. In this section, we briefly discuss the aspects of the compiler that are relevant for this article.

@@

\section{Performance evaluation}
\label{sec:perf-results}

@@

The Zenodo system was used to archive all packages used to perform the experiments: Firedrake \cite{lawrence_mitchell_2016_49284}, PETSc \cite{barry_smith_2016_49285}, petsc4py \cite{firedrake_2016_49283}, FIAT \cite{marie_e_rognes_2016_49280}, UFL \cite{martin_sandve_alnaes_2016_49282}, FFC \cite{anders_logg_2016_49276}, PyOP2 \cite{florian_rathgeber_2016_49281} and COFFEE \cite{fabio_luporini_2016_49279}. The experiments can be reproduced using a publicly available benchmark suite~\cite{florian_rathgeber_2016_49290}. Further, the generated code is available at \cite{generated_code}.

@@

Following the methodology adopted in~\citeN{quadrature1}, we vary the following parameters:
\begin{itemize}
\item the polynomial degree of test, trial, and coefficient (or ``pre-multiplying'') functions, $q \in \lbrace1, 2, 3, 4\rbrace$
\item the number of coefficient functions $\mathrm{nf} \in \lbrace0, 1, 2, 3\rbrace$
\end{itemize}
While constants of our study are
\begin{itemize}
\item the test, trial, and coefficient functions belong to the space of (possibly vector-valued) continuous piecewise-defined polynomials of degree $q$
\item the mesh: tetrahedral with a total of 4374 elements
\item exact numerical quadrature (we employ the same scheme used in~\citeN{quadrature1}, based on the Gauss-Legendre-Jacobi rule)
\item double-precision arithmetic
\end{itemize}

@@

The memory transfer has sometimes a non-negligible impact, especially when low order polynomials are used. This is more evident in the simpler bilinear forms (e.g., \texttt{Mass}), when the speed-ups over the baseline are limited. The effect of the memory transfer is quantified, to some extent, in the sections below. For a comprehensive analysis of its cost, the reader is invited to refer to~\cite{extrusion-paper}, a study based upon the same Firedrake technology used in this article.

\subsection{Autovectorization}
We chose the mesh dimension and the function spaces such that the inner loop sizes would always be a multiple of the machine vector length. This ensured autovectorization in the majority of code variants\footnote{We verified the vectorization of inner loops by looking at both compiler reports and assembly code.}. The biggest exception is \texttt{quad}, due to the presence of indirection arrays in the generated code. In \texttt{tens}, loop nests are fully unrolled, so the standard loop vectorization is not feasible; the compiler reports suggest, however, that block vectorization \cite{SLP-vect} is often triggered. In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have identical structure, with loop vectorization being regularly applied.

@@

We first notice the lack of improvements when $q=1$. The operation count of \texttt{cfO2} and \texttt{tensor} is roughly 9$\times$ lower than that of the alternatives (tens instead of hundreds operations), but nonetheless no gain in execution time is obtained. This is due to the fact that local assembly is memory-bound. For instance, in the simplest case $nf=0$, in addition to the determinant of the Jacobian inverse, \texttt{tens} only requires three unique multiplications to evaluate the local element matrix (26 operations in total), whereas almost a hundred memory accesses are performed. In particular, we have: 12 loads for the vertex coordinates; for each vertex, 1 load to indirectly access its data; for each of the 16 local elements, 4 loads (optimistically) and 1 store to update an entry in the global matrix in compressed sparse row format. This unsurprisingly leads to a very low arithmetic intensity (i.e., $\frac{\text{operation count}}{\text{bytes moved}}$), which collocates the kernels in a memory-bound regime.

@@

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.685]{perf-results/elasticity}}
\caption{Performance evaluation for the bilinear form arising in an \textit{elastic} model. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
\end{figure}

@@

\subsection{On the correlation between operation count and performance}
As reiterated throughout the paper, the execution time of a kernel is influenced not only by the operation count, but also by the chosen code generation strategy (e.g., loop unrolling rather than preserving the loop structure). Figure~\ref{fig:flops-correlation} aggregates the \texttt{cfO2} results from the previous sections and displays them as a function of the reduction in executed flops. Besides summarizing the performance improvements achieved in this article, this figure highlights a number of aspects. The speed-ups achieved by \texttt{cfO2} tend to increase with the complexity of the operators and $q$. Further, the pre-multiplying functions have a limited impact in complex operators, and they occasionally have a negative effect in simple and small problems. Finally, this plot shows that even a powerful general-purpose compiler like Intel's is inadequate to fully exploit the structure of non-trivial finite element operators (analogous behaviors were observed with different compilers).

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.55]{perf-results/overview}}
\caption{Relationship between operation counts and performance achieved by \texttt{cfO2}.}
\label{fig:flops-correlation}
\end{figure}

@@

The transformations and the cost models described in this paper assume that a local assembly kernel is executed sequentially by a single process for each mesh element. In Firedrake, parallelism is always achieved at the mesh level, by distributing element partitions to different processes. However, this strategy could be revisited, for instance, in the case of high order methods (not treated in this article), when the working set to execute an element does not fit in L3 cache. This would require parallelism at the level of the basis functions loops. The cost models should also be extended, since certain rewrite operators (e.g., factorization and generalized code motion) could on one hand reduce the operation count, but at the same time destroy otherwise easily parallelizable computation.
